### 2009-06-04 s4.tex
renamed software appendix (Text S3) to Text S4; added new Text S3 for TKFST grammars; added a bunch of other stuff

% Title must be 150 words or less
\begin{flushleft}
  {\Large
    \textbf{\titlestring: \supptext}
  }
\\
\authorstring
\end{flushleft}

\newpage
\tableofcontents

\newpage
\section{Exact elimination of SCFG null cycles}

The following section describes how to transform any SCFG so as to remove null cycles while preserving access to the full posterior probability distribution over parses, including parses with null cycles.

\subsection{Definitions}

Let $G = ( \Omega, \gnons, \grules, P )$ be a stochastic context-free grammar (SCFG)
consisting of
a set of terminal symbols $\Omega$,
a set of nonterminal symbols (a.k.a. ``states'') $\gnons$,
a set $\grules$ of {\em production rules} $L \to R$ (where $L \in \gnons$ and $R \in \grhs$)
and a probability function on the rules, $P:(\gnons \times \grhs) \to [0,\infty)$.

% (The rule set $\grules$ can be considered redundant to the specification of the grammar, in that $\rho \notin \grules \Leftrightarrow P(\rho)=0$.
% The interpretation is that $P$ is a probabilistic weight function and $\grules$ is informally the set of rules with non-zero weight.)

Let $\leftside{A}$ be the set of rules that can be applied to nonterminal $A \in \gnons$ (i.e., rules in which $A$ appears on the left),
and let $\rightside{A}$ be the set of rules that can generate nonterminal $A$ (i.e.,rules in which $A$ appears on the right, including bifurcations which also generate another nonterminal as well as $A$).
Let $\allrules{A} = \leftside{A} \cup \rightside{A}$ be the set of all rules involving $A$.
Define these also on sets of nonterminals, e.g., $\leftside{\gnulls} = \bigcup_{A \in \gnulls} \leftside{A}$ for $\gnulls \subseteq \gnons$.

Let $\gtrees$ be the set of all parse trees for $G$.
Suppose that parse tree $T \in \gtrees$ makes $n_T(\rho)$ uses of rule $\rho$;
then define the parse tree likelihood $P(T) = \prod_{\rho \in \grules} \left[ P(\rho) \right]^{n_T(\rho)}$.

If $\sum_{T \in \gtrees} P(T) = 1$, we say that $G$ is {\em probabilistically normalized by parse tree}.
If $\sum_{\rho \in \leftside{A}} P(\rho) = 1$ for all $A \in \gnons$, we say that $G$ is {\em probabilistically normalized by production rule}.
Note that normalization by production rule $\Rightarrow$ normalization by parse tree.

Let $S \in \Omega^\ast$ denote a terminal sequence.
Let $\mathrm{seq}:\gtrees \to \Omega^\ast$ be the function mapping a parse tree to its terminal sequence.
and let $\mathrm{root}:\gtrees \to \gnons$ be the function returning the root nonterminal of a parse tree.
The {\em inside probability of $S$ rooted at $A$} is $P(S|A)\ = \sum_{T: \mathrm{seq}(T) = S,\mathrm{root}(T) = A} P(T)$.
Of particular relevance to null state elimination is the probability $P(\epsilon|A)$,
where $\epsilon$ is the empty sequence.
This is the probability that a nonterminal $A$ will expire without generating any sequence.

Following earlier formalisms \cite{Durbin98,HolmesRubin2002a,Holmes2005},
we say that the grammar $G$ has {\em RNA normal-form rules} if
each rule in $\grules$ takes one of the following four forms:

\noindent \hrulefill
\begin{description}
\item[Termination rules] with one nonterminal on the left and the empty string on the right
\begin{eqnarray*}
A & \to & \epsilon
\end{eqnarray*}
\item[Transition rules] with one nonterminal on the left and the right
\begin{eqnarray*}
A & \to & B
\end{eqnarray*}
\item[Bifurcation rules] with one nonterminal on the left and two on the right
\begin{eqnarray*}
A & \to & B\ C
\end{eqnarray*}
\item[Emission rules] with one nonterminal on the left and right and at least one terminal on the right
\begin{eqnarray*}
A & \to & x\ B \\
A & \to & B\ y \\
A & \to & x\ B\ y
\end{eqnarray*}
\end{description}
\noindent \hrulefill

\noindent
Here $A,B,C \in \gnons$ and $x,y \in \Omega$.

Suppose that grammar $G$ has RNA normal-form rules.
We further say that $G$ has {\em RNA normal-form states} if
each nonterminal (state) $A \in \gnons$ takes one of the following forms:

\noindent \hrulefill
\begin{description}
\item[Null states]: $\leftside{A}$ contains only transition and termination rules.
\item[Bifurcation states]: $\leftside{A}$ contains exactly one bifurcation rule, $A \to X\ Y$, where $X \neq Y$ and $X,Y$ are both null states.
\item[Emit states]: $\leftside{A}$ contains only emission rules. Further, $\leftside{A} = \rightside{B}$ for some null state $B$.
This null state $B$ is called $A$'s {\em post-emit} state.
\end{description}
\noindent \hrulefill

Define a {\em null cycle} to be a nonterminal $A$
and a sequence of rules $\rho_1, \rho_2 \ldots \rho_k$ that, when applied consecutively to $A$, leave $A$ unchanged;
that is, $\rho_k(\rho_{k-1} \ldots \rho_2(\rho_1(A))) = A$.
Define a {\em null subtree} to be a null cycle using at least one bifurcation rule.

Suppose that $G = ( \Omega, \gnons, \grules, P )$ and $G' = ( \Omega, \gnons', \grules', P' )$ are two grammars.
We say that $G$ and $G'$ are {\em equivalent in sequence} if there is a mapping between nonterminals, $f:\gnons \to \gnons'$,
such that $P(S|A) = P(S|f(A))$.
We say that $G$ and $G'$ are {\em equivalent in parse} if there is a mapping between parse trees, $g:\gtrees \to \gtrees'$,
such that $P'(T') = \sum_{T:T'=g(T)} P(T)$ and we can define
\[
P(T|T') = P(T|g(T)=T') = \frac{P(T)}{P'(T')}
\]

Note that equivalence in parse $\Rightarrow$ equivalence in sequence.

Suppose that a grammar $G$ contains null cycles.
We seek to transform $G$ into a grammar $G''$ that is equivalent in parse and sequence, but has no null cycles.
If $G$ is normalized by parse tree, then $G''$ will be too; but $G''$ is not (necessarily) normalized by production rule.

We do the transformation in two steps, $G \to G' \to G''$.
We also provide a stochastic ``null cycle restoration'' algorithm for sampling from $P(T|T'')$.

\subsection{Eliminating null bifurcations}

Note first that any SCFG can be transformed into an equivalent one with RNA normal-form states by adding null states.
Without loss of generality, we therefore consider SCFGs with RNA normal-form states.

Let $G = ( \Omega, \gnons, \grules, P )$ be such an SCFG.
Let $\gnulls \subseteq \gnons$ be the set of null states in $\gnons$, {\bf excluding} post-emit states.
Let $\gemits \subset \gnons$ be the set of post-emit states.
Let $\gbifs \subseteq \gnons$ be the set of bifurcation states.

We here define the grammar $G'$,
\[
G' = \left( \Omega, \gnons', \grules', P' \right)
\]
which is equivalent to $G$ in sequence but has no null subtrees.

The new states $\gnons'$ are defined as follows.
Start with $\gnons$; remove bifurcation states;
then introduce a new state $N'$ for every null $N \in \gnulls$ and three new states $B^0, B', B^2$ for every bifurcation $B \in \gbifs$.

\[
\gnons' = 
\left( \bigcup_{N \in \gnulls} \{ N' \} \right) \cup \left( \bigcup_{B \in \gbifs} \{ B^0, B', B^2 \} \right) \cup \gnons \setminus \gbifs
\]

The idea is that $N'$ (in $G'$) is a null state that is equivalent to null state $N$ (in $G$) for non-empty sequence only.
That is, for any terminal sequence $S$,

\[
P'(S|N') = \left\{ \begin{array}{ll} P(S|N) & \mbox{if $S \neq \epsilon$} \\ 0 & \mbox{if $S = \epsilon$} \end{array} \right.
\]

Similarly, $B'$ is a null state that is equivalent to bifurcation state $B$ for non-empty sequence,
while $B^2$ is a null state that is is equivalent to $B$ for parse trees where both children of $B$ are non-empty.

In contrast, $B^0$ is {\bf exactly} equivalent to $B$ in sequence.
However, null subtrees in $B^0$ are explicitly accounted for,
their probabilities factored into the transitions $B' \to L'$ and $B' \to R'$ and the termination $B^0 \to \epsilon$,
using the inside probabilities for empty sequences, $P(\epsilon|L)$ and $P(\epsilon|R)$.

The probabilities $P(\epsilon|X)$ are related by the following system of equations

\[
P(\epsilon|X) = P(X \to \epsilon) + \sum_Y P(X \to Y) P(\epsilon|Y) + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R)
\]

which are nonlinearly coupled (via bifurcations) but may be solved numerically, e.g., by the Newton-Raphson method,
or by iterated approximation starting from a lower bound $P(\epsilon|X) \geq 0$.

The new rules and their probabilities are
\[
\begin{array}{rrcl}
\forall \rho \in (\grules \setminus \allrules{\gbifs}): & P'(\rho)             & = & P(\rho) \\
                                \forall N \in \gnulls: & P'(N' \to \epsilon)  & = & 0  \\
                          \ldots \forall B \in \gbifs: & P'(N' \to B')        & = & P(N \to B) \\
                         \ldots \forall M \in \gnulls: & P'(N' \to M')        & = & P(N \to M) \\
                         \ldots \forall E \in \gemits: & P'(N' \to E)         & = & P(N \to E) \\
               \forall (A \to B) \in \rightside{\gbifs}: & P'(A \to B^0)        & = & P(A \to B) \\
              \forall (B \to LR) \in \leftside{\gbifs}: & P'(B^0 \to \epsilon) & = & P(B \to LR)\ P(\epsilon | L)\ P(\epsilon | R) \\
                                                       & P'(B^0 \to B')       & = & 1 \\
                                                       & P'(B' \to \epsilon)  & = & 0 \\
                                                       & P'(B' \to L')        & = & P(B \to LR)\ P(\epsilon | R) \\
                                                       & P'(B' \to R')        & = & P(B \to LR)\ P(\epsilon | L) \\
                                                       & P'(B' \to B^2)       & = & 1 \\
                                                       & P'(B^2 \to L'\ R')   & = & P(B \to LR) \\
\end{array}
\]

Note that grammar $G'$, like $G$, has RNA normal-form states.
Note also that $G'$ is not, in general, normalized by production rule;
however, $G'$ is equivalent to $G$ in parse and is therefore normalized by parse tree (if $G$ is).

\subsection{Eliminating null states}

We now proceed to eliminate null cycles from $G'$.
Since null subtrees have been eliminated, the remaining null cycles use only transition rules.

We create the grammar
\[
G'' = \left( \Omega, \gnons', \grules'', P'' \right)
\]
which has the same nonterminals as $G'$, but different rules and rule probabilities.

Let $\gnulls' \subseteq \gnons'$ be the set of null states in $\gnons'$, {\bf including} post-emit states.

Define ${\bf t}$, the {\em transition matrix} of $G'$, as $t_{XY} = P'(X \to Y)$ for all $X,Y \in \gnons'$.
The effective transition probability $q_{XY}$ between two states $X,Y$
sums over all paths through null states (in the summand, $n$ is the length of the null path):
\[
{\bf q} = \sum_{n=0}^\infty {\bf t}^n = ({\bf 1}-{\bf t})^{-1}
\]
Here ${\bf 1}$ is the identity matrix.
The matrix inverse may be computed in the usual ways (Gauss-Jordan elimination, LU decomposition, etc.)

Since the bifurcation states of $G'$ explicitly generate non-empty sequence,
the system of equations relating the probabilities $P'(\epsilon|X)$ is completely linear
and may be solved in the same way.
Let $u_X = P'(\epsilon|X)$ and $v_X = P'(X \to \epsilon)$.
Then

\[
{\bf u} = {\bf v} + {\bf tu}
\]

whose solution is ${\bf u} = {\bf qv}$.
Thus $P'(\epsilon|X) = \sum_Y q_{XY} P'(Y \to \epsilon)$.

We now define $P''$ as follows.
\begin{itemize}
\item For all rules $\rho \in \grules'$ that are {\bf not} transitions or terminations, set $P''(\rho) = P'(\rho)$.
\item For terminations from null states $X \in \gnulls'$, set $P''(X \to \epsilon) = u_X$.
(NB this may create some terminations $X \to \epsilon$ that were not present in $G'$.)
\item For transitions from null states $X \in \gnulls'$ to non-null states $Y \notin \gnulls'$, set $P''(X \to Y) = q_{XY}$.
(NB this may create some transitions $X \to Y$ that were not present in $G'$.)
\item For transitions {\bf to} null states $X \in \gnulls'$, set $P''(A \to X) = 0$.
\end{itemize}

Although we have left null states in $G''$, they now have no incoming transitions
and are inaccessible unless they are post-bifurcation states (i.e.,states which appear on the right-hand side of bifurcation rules), post-emit states, or the start (root) state.
All other null states can therefore be dropped from $\gnons''$.

\subsection{Restoring null states}

Suppose that $\rho''$ is a rule in $G''$.
Algorithm~\ref{alg:restoreTransitions} samples from the distribution of equivalent parse subtrees in $G'$ (possibly containing null cycles).
In order to sample from $P(T'|T'')$ we need simply map Algorithm~\ref{alg:restoreTransitions} to each rule in $T''$.

For parameter estimation by Expectation Maximization and some other applications, it is useful to know the expected number
of times that a transition was used, summed over the posterior distribution of parse trees (including those with null cycles).
If $n''(\rho'')$ is the expected number of times that transition $\rho''$ was used according to an Inside-Outside computation on $G''$,
then the corresponding expectations $n'(\rho')$ are given by

\begin{eqnarray*}
       n'(X \to Y) & = & n''(X \to Y) \frac{P''(X \to Y)}{q_{XY}} + \sum_Z n''(X \to Z) \frac{P''(X \to Y) q_{YZ}}{q_{XZ}} \\
n'(X \to \epsilon) & = & n''(X \to \epsilon) \frac{P''(X \to \epsilon)}{P''(\epsilon | X)}
 + \sum_W n''(W \to \epsilon) \frac{q_{WX} P''(X \to \epsilon)}{P''(\epsilon | W)}
\end{eqnarray*}

Expectations for other rules (bifurcations and emissions) are the same for $G'$ as $G''$.

\subsection{Restoring null bifurcations}

Suppose that $\rho'$ is a rule in $G'$.
Algorithm~\ref{alg:restoreBifurcations} samples from the distribution of equivalent parse subtrees in $G$ (possibly containing null subtrees).
This algorithm also calls Algorithm~\ref{alg:sampleNullSubtree}, which samples from the distribution of empty subtrees rooted at a particular nonterminal.
In order to sample from $P(T|T')$ we need simply map Algorithm~\ref{alg:restoreBifurcations} to each rule in $T'$.

If $n'(\rho')$ is the expected number of times that rule $\rho'$ was used by $G'$,
then the corresponding expectations $n(\rho)$ are given by

\begin{eqnarray*}
    n(B \to L\ R) & = & n'(B' \to L') + n'(B' \to R') + n'(B^0 \to L'\ R') + d(B \to L\ R) \\
       n(X \to Y) & = & n'(X \to Y) + d(X \to Y) \\
n(X \to \epsilon) & = & n'(X \to \epsilon) + d(X \to \epsilon)
\end{eqnarray*}
Expectations for emissions are the same for $G$ as $G'$.
In the above expressions $d(\rho)$ is the expected usage of rule $\rho$ by null subtrees:

\[
d(\rho) = \sum_{(B \to LR) \in \grules} \left[
n'(B' \to L') c_R(\rho)
+ n'(B' \to R') c_L(\rho)
+ n'(B^0 \to \epsilon) (c_L(\rho) + c_R(\rho))
\right]
\]

where $c_W(\rho)$ is the expected usage of rule $\rho$ by an empty parse tree rooted at $W$,
given by

\begin{eqnarray*}
c_X(\rho) P(\epsilon | X) & = & P(X \to \epsilon) \delta_{\rho=(X \to \epsilon)}
 + \sum_Y P(X \to Y) P(\epsilon|Y) \left( c_Y(\rho) + \delta_{\rho=(X \to Y)} \right)
\\ & &
 + \sum_{L,R} P(X \to LR) P(\epsilon|L) P(\epsilon|R) \left( c_L(\rho) + c_R(\rho) + \delta_{\rho=(X \to LR)} \right)
\end{eqnarray*}

where $\delta_U$ is the Kronecker delta (1 if condition $U$ is true, 0 if it is false).
Note that in contrast to the system of equations for $P(\epsilon|X)$, this is a linear system of equations
of the form ${\bf c} = {\bf Mc} + {\bf k}$, which may be solved by matrix inversion:
${\bf c} = ({\bf 1}-{\bf M})^{-1} {\bf k}$.

\newpage
\section{Two-sequence transducer models}

We here give formal definitions of two-sequence models and the state machines which
generate the factored probability distribution $P (X, Y | \Delta T) = P (X) \cdot P (Y | X, \Delta T)$,
where the marginal $P(X)$ is generated by a {\bf singlet transducer}
and the conditional $P (Y | X, \Delta T)$ by a {\bf branch transducer}.
We refer to the branch transducer as $\theta$, so the conditional distribution
is more precisely $P (Y | X, \Delta T, \theta)$.

\subsection{Formal grammars} \seclabel{grammars}

There is a close relationship between formal grammars and the singlet and branch transducer abstract state machines.
By labeling the nonterminals of a regular or stochastic context-free grammar (SCFG) as states of an abstract machine, allowing these states to absorb and emit 
appropriate terminal symbols,\footnote{The relationship between formal grammars and abstract state machines is more transparent
in the Mealy machine view, but the Moore machine view turns out to be more useful for our purposes.},
and carefully assigning transition and emission weights,
we can create a state machine which generates the same language, with the same distribution of weights, as that produced by the original grammar.\footnote{We speak of
weights associated with a transition rather than probabilities in order to allow for more general models.}
We therefore use the terms ``nonterminal'' and ``state'' interchangeably
and refer to the ``parse tree'' generated by singlet and branch transducers.

Phrased more precisely, there is an isomorphism between the singlet and branch transducers of two-sequence models
and Pair SCFGs.  Practically speaking, this means that for every joint distribution $P(X, Y | \Delta T)$ 
generated by a singlet and branch transducers, there exists a Pair SCFG which generates the same distribution.

\subsection{States, transitions and emissions}
A singlet transducer has states of type $\Sstart$ and $\Sinsert$.
Each state $\phi\in\Phi$ of the branch transducer has type
\[ \type(\phi) \in \lbrace\Sstart,\,\Sinsert,\,\Smatch,\,\Swait,\,\Send\rbrace \, . \]
State typing is as follows:
\begin{itemize}
  \item A $\Sstart$ state begins a branch of the parse tree.
  \item An $\Sinsert$ state emits, but does not absorb, symbols.
  \item A $\Smatch$ state absorbs and (possibly) emits symbols.
  \item A $\Swait$ state is a null state which allows a branch transducer to ``pause''
    while it waits for an input symbol from another transducer.
    In the two-sequence case, the input symbols is emitted by the singlet transducer generating
    the ancestral sequence.
  \item An $\Send$ state ends a branch of the parse tree.
\end{itemize}
The names of state types are similar to the $\{\Sstart,\,\Sinsert,\,\Smatch,\,\Send\}$ states of the familiar Pair HMM.
Deletions are handled as a special case by states of type $\Smatch$ which absorb but do not emit symbols pairs.
Only states of type $\Smatch$ can absorb symbol pairs $(x,y)$.

Bifurcations in the grammar, when considered as emission of nonterminals, can be handled analogously
to terminal emission.  States $\phi: \type(\phi) \in \{\Sinsert,\,\Smatch\}$ can emit nonterminal pairs $(c\,d)$,
where $c$ or $d$ can be the null symbol, and the emitted pairs $(c\,d)$ can be absorbed by states of type $\Smatch$.

The emission weight of a nonterminal pair $(c\,d)$ from the state $b: \type(b) = \Smatch$, 
conditional on absorption of a nonterminal pair $(l\,m)$, is 
\[ e_b ( c\,d|l\,m,\theta ) \, . \]

The functions $\emit()$ and $\absorb()$ are defined to return the emission or absorption of a particular state,
\begin{align*}
  \emit(\phi) &:= \left\{\begin{array}{ll} (x,y) & x,y = \mathrm{terminal\,\,or\,\,null} \\(c\,d) & c,d = \mathrm{nonterminal\,\,or\,\,null}. \end{array} \right. \\
  \absorb(\phi) &:= \left\{\begin{array}{ll} (x,y) & x,y = \mathrm{terminal\,\,or\,\,null} \\(c\,d) & c,d = \mathrm{nonterminal\,\,or\,\,null}. \end{array} \right.
\end{align*}

We frequently use the notation ${}^{uv}\phi^{xy}$ to indicate that a state of type $\Smatch$ absorbs a symbol pair $(u,v)$
and emits a pair $(x,y)$.
The notation for bifurcations is slightly different:
A bifurcation state which left-emits a nonterminal $d$ and makes a transition to a state $\phi$ with weight 1 is written as $\bifurc{B}{d}{\phi}$.

A transition between states $a$ and $b$ of the branch transducer $\theta$ has a weight
\[ t(a,b|\theta) = t(a \to b|\theta). \]
Terminal emission is handled by states $\phi: \type(\phi) \in \{\Smatch,\,\Sinsert\}$, which emit symbol pairs $(x,y)$,
where $x$ or $y$ can be the null symbol.
In this paired-emission perspective, left single-terminal emissions $x$ are represented as $(x\,\Tnull)$;
right single-terminal emissions are handled similarly.
The weight of an emission of a terminal pair $(x,y)$ from a state $b: \type(b) = \Smatch$, 
conditioned on absorption of a terminal pair $(u,v)$, is 
\[ e_b ( x,y|u,v,\theta ) \, . \]
Recall that the emission weights of states of type $\Smatch$ are defined conditioned upon the absorbed symbols.

\newpage
\section{Multiple-sequence transducer models} \seclabel{evolutionarymodels}

We can use our two-sequence models to construct a model of many sequence related by a
guide phylogenetic tree.
The guide tree specifies the (conjectured) phylogenetic relationship of all sequences.
A singlet transducer, which emits, but does not absorb, symbols, lies at the root of the guide tree and 
serves as a generative model of the ancestral sequence.
A branch transducer represents the evolution of an ancestral sequence into a single descendant 
sequence, that is, the action of evolution along the single-branch tree (Ancestor $\to$ Descendant).
To represent the evolution of an ancestral sequence into many descendant sequences
(whose phylogenetic relationship is specified by the guide tree),
we place a branch transducer on each branch of the guide tree to create a multiple-sequence model.

If the branch grammar has no bifurcations and only left or right emissions are allowed, then
the language generated is a regular string language and the corresponding jointly normalized abstract state machine is 
an HMM.  In this simplest case our formalism for creating a multiple-sequence model reduces to that given 
by \cite{Holmes2003} for combining HMMs on a guide tree.

\subsection{The guide tree}
The nodes of the tree are labeled $1,\dots,N$ in the order reached by any preorder depth-first traversal of the tree.
The length of each branch $(\parent(m) \to m)$ is given by the evolutionary time $t_m$.
To specify ancestor-descendant relationships, we introduce notation:
$\mDn$ ($\mNDn$) means node $m$ is descended from (not descended from) node $n$, and
$\mDEn$ ($\mNDEn$) means node $m$ is descended from or identical to (not descended from and not identical to) node $n$.

\subsection{States, transitions and emissions}
The multiple-sequence model is formed by the composition/intersection of $(N-1)$ branch transducers 
such that there is a branch transducer on each branch $(\parent(m) \to m);\, m = 2,\dots,N$ of the guide tree and a singlet transducer at the root node.
Our framework allows for the placement of different branch transducers, with a unique set of nonterminals (state space) $\Phi$ and
allowed transitions between states and corresponding weights, on each branch.  $\theta^{(m)}$ denotes 
the branch transducer governing evolution along the branch $(\parent(m) \to m)$ of the guide tree.

States of the multiple-sequence model are represented by as N-dimensional vectors $\bvec{a}$,
\[ \bvec{a} = \statevec{a} \, . \]
These states are typed as
\[ \type(\bvec{a}) \in \lbrace\Sstart,\,\Semit,\,\Sbifurc,\,\Snull,\,\Send\rbrace \, . \]
State typing is as follows:
\begin{itemize}
  \item A $\Sstart$ state begins a branch of the parse tree of one or more of the $N$ sequences on the phylogenetic tree.
  \item An $\Semit$ state emits symbols (terminals) to one or more of the $N$ sequences.
  \item A $\Sbifurc$ state corresponds to a bifurcation in the parse tree of
    one or more of the $N$ sequences.
  \item A $\Snull$ state corresponds to any non-$\Send$ state which represents neither
    an emission or bifurcation.
  \item An $\Send$ state ends a branch of the parse tree.
\end{itemize}
States are typed according to the transition by which they are reachable
(details of the typing are given in \secref{transitions}).

Transitions and emissions of the multiple-sequence model are defined in terms of the transitions and emissions 
of the branch transducers at each node as well as the singlet transducer at the root node.
The weight of a transition $\bvec{a} \to \bvec{b}$ is therefore
\beqn
t(\bvec{a},\bvec{b}) = \prod_{m|a_m \ne b_m} \transweight{m}{a_m}{b_m} \, ,
\eeqn
and the weight of an emission $(\bvec{x}\,\bvec{y})$ from a state $\bvec{b}$ is
\beqn
\totalemissionweight{\bvec{b}}{\bvec{x}\,\bvec{y}} = \prod_m e_{b_m}\left(x_m\,y_m|x_{\parent(m)}\,y_{\parent(m)},\theta^{(m)}\right) \, .
\eeqn
We frequently will not explicitly write out the conditional dependence on the absorbed terminals $(x_{\parent(m)}\,y_{\parent(m)})$,
but the reader should keep in mind that in general emission weights will depend on the absorbed symbols.

\subsection{Formal grammars} \seclabel{multigrammars}

Analogously to the case with two-sequence models (\secref{grammars}),
there exists a one-to-one mapping between the multiple-sequence models generated by our model-construction algorithm 
and multi-sequence SCFGs.
In other words, given a singlet and branch transducers of a two-sequence model, as well
as a guide tree relating the extant sequences, there exists a corresponding multi-sequence SCFG
which generates the same joint probability distribution $P(X_1 , ... , X_n)$.

\subsection{Constructing the state graph} \seclabel{stategraph}

As described in the paper,
we need a way to efficiently construct the state graph of the multi-sequence model,
where the state graph consists of a list of accessible states and the possible transitions between them.
This state graph can be constructed by an uninformed depth-first search,
where at each step of the search we obtain the possible child nodes 
by applying one of the following possible transitions of the multi-sequence model:
\begin{enumerate}
\item {\bf Null transition}:
  The state of a single branch transducer is updated, with no terminal emission or bifurcation.
\item {\bf Terminal emission}:
  A state makes a transition to a $\Sinsert$ state.  The emitted symbol is passed down the guide
  tree to all descendant branch transducers, which transition to states of type $\Smatch$.
\item {\bf Bifurcation}:
  A state makes a transition to a special $\Sinsert$ state which emits a new branch of the parse tree.
  The emitted symbol is passed down the guide tree to all descendant branch transducers, 
  which transition to states of type $\Smatch$.
\item {\bf End transition}:
  The singlet transducer associated with the root sequence can transition to the $\Send$ state,
  signaling that this parse tree is finished.
\end{enumerate}

Each of these transitions is explained in detail in the following section.

\subsection{Allowed transitions} \seclabel{transitions}

Following \cite{Holmes2003},  we let transitions of the multi-sequence model begin at the active node and cascade down the guide tree
as appropriate.  Unless defined otherwise, node $n$ is the active node of the multi-sequence model with state $\bvec{a}$, 
\begin{align}
  n &= \activenode{a} \\
  &= \mathrm{argmax}_m \left\{\type(a_m) \notin \{\Swait,\,\Send\}\right\} \, . \label{eqn:activenode} 
\end{align}
Each possible allowed transition is obtained by making a valid change (updating) the 
state of the singlet or one or more of the branch transducers of the multi-sequence model.

\subsubsection*{Null Transitions}

\[ \statevec{a} \to \statevec{b} \]

\begin{align} \label{eqn:nulltransition}
  \type(\bvec{b}) &= \Snull  \\
  \weight(\bvec{a}\to\bvec{b}) &= t(\bvec{a},\bvec{b}) \\
  &= \transweight{n}{a_n}{b_n} .
\end{align}

This transition updates the state of the branch transducer at the active node $n$ of the guide tree, leaving 
the rest unchanged, with no corresponding terminal emission or bifurcation in the grammar.
Nodes other than the active node do not change state, $a_m = b_m \,\forall\, m \ne n$, and the only 
allowable transitions of this form are to states $b_n:\,\type(b_n) \in \{\Sstart,\,\Swait\}$.  Transitions
to states of type $\Sinsert$ or $\Smatch$ result in emissions, and transitions to the end state $\Send$
are handled as a special case.

\subsubsection*{Terminal Emission} \seclabel{terminalemission}
\[ \statevec{a} \to \statevec{x}\,\statevec{b}\,\statevec{y} \]

\begin{align} \label{eqn:terminalemission}
  \type(\bvec{b}) &= \Semit \\
  \weight(\bvec{a}\to\bvec{x}\,\bvec{b}\,\bvec{y}) &=  t(\bvec{a},\bvec{b}) \cdot \totalemissionweight{\bvec{b}}{\bvec{x}\,\bvec{y}} \\
  &= \left[ \prod_{m|a_m \ne b_m} \transweight{m}{a_m}{b_m} \right] \cdot \left[ \prod_m \emissionweight{m}{b_m}{x_m\,y_m} \right] ,
\end{align}
where we are defining states with no emissions to emit $(\Tnull\,\Tnull)$ with weight 1, 
$\emissionweight{m}{b_m}{x_m\,y_m} = 1$ if $(x_m\,y_m) = (\Tnull\,\Tnull)$ and $\emit(b_m) = \Tnull$.

The active node $n$ makes a transition to a state $b_n:\,\type(b_n) = \Sinsert$, emitting
a terminal symbol pair $\emit(b_n) = (x_n\,y_n)$.  This symbol pair is passed down the guide tree to descendant nodes $\{m|\mDn, \type(a_m) \ne \Send\}$, 
forcing them to make a transition from states of type $\Swait$ to states of type $\Smatch$, which can absorb
terminal pairs $(x,y)$.  Qualitatively, this transition and emission could represent the evolution
of two paired nucleotides along the subtree rooted at node $n$ of the complete guide tree.
If one of $\bvec{x}$ or $\bvec{y}$ is null, then \eqref{eqn:terminalemission} could represent the evolution of
a single unpaired nucleotide along the subtree rooted at node $n$.

\paragraph{Left emission.}
All terminals $\bvec{y}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission} are null.

\begin{tabular}{rl}
  Nodes $\mNDEn$: & $a_m = b_m$. \\
  & $(x_m\,y_m) = (\Tgap\,\Tnull)$. \\
  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
  & $(x_n\,y_n) = \emit(b_n)$. \\
  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (x_{\parent(m)}\,\Tnull)$. \\
  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tgap\,\Tnull) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull \end{array} \right. $
\end{tabular}

\paragraph{Right emission.}
All terminals $\bvec{x}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission} are null.

\begin{tabular}{rl}
  Nodes $\mNDEn$: & $a_m = b_m$. \\
  & $(x_m\,y_m) = (\Tnull\,\Tgap)$. \\
  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
  & $(x_n\,y_n) = \emit(b_n)$. \\
  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (\Tnull\,y_{\parent(m)})$. \\
  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tnull\,\Tgap) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull . \end{array} \right. $
\end{tabular}

\paragraph{Paired emission.}

There is at least one non-null terminal in both $\bvec{x}$ and $\bvec{y}$ in the transition $\bvec{a} \to \bvec{x}\,\bvec{b}\,\bvec{y}$ \eqref{eqn:terminalemission}.

\begin{tabular}{rl}
  Nodes $\mNDEn$: & $a_m = b_m$. \\
  & $(x_m\,y_m) = (\Tgap\,\Tgap)$. \\
  Node $n$: & $b_n:\,\type(b_n) = \Sinsert,\,\exists$ a transition $a_n \to b_n$ in the branch transducer $\theta^{(n)}$. \\
  & $(x_n\,y_n) = \emit(b_n)$. \\
  Nodes $\mDn$: & Either $a_m = b_m$, $\emit(b_{\parent(m)}) = \Tnull$ or $\type(a_m) = \Send$ \\
  & $\quad$ or $\emit(b_{\parent(m)}) = \absorb(b_m) = (x_{\parent(m)}\,y_{\parent(m)})$. \\
  & $(x_m\,y_m) = \left\{ \begin{array}{ll} (\Tgap\,\Tgap) & \emit(b_m) = \Tnull \\ \emit(b_m) & \emit(b_m) \ne \Tnull . \end{array} \right. $
\end{tabular}

\subsubsection*{Bifurcations}
\[ \statevec{a} \to \statevec{c}\,\statevec{b}\,\statevec{d} \]

\begin{align} \label{eqn:nonterminalemission}
  \type(\bvec{b}) &= \Sbifurc \\
  \weight(\bvec{a}\to\bvec{c}\,\bvec{b}\,\bvec{d}) &=  t(\bvec{a},\bvec{b}) \cdot \totalemissionweight{\bvec{b}}{\bvec{c}\,\bvec{d}} \\
  &= \left[ \prod_{m|\,a_m \ne b_m} \transweight{m}{a_m}{b_m} \right] \cdot \left[ \prod_m \emissionweight{m}{b_m}{c_m\,d_m} \right] .
\end{align}
where we are defining the emission weight of the $\Send$ nonterminal to be 1, $\emissionweight{m}{b_m}{c_m\,d_m} = 1$ if $c_m = \Send$ or $d_m = \Send$.

>>>>

% Title must be 150 words or less
\begin{flushleft}
  {\Large
    \textbf{\titlestring: \supptext{4}}
  }
\\
\authorstring
\end{flushleft}
